(I have decided to skip this)

## The honorable mention `multidplyr`


The package `multidplyr` is tidyverse solution to multiprocess data management. However, it is not released to CRAN yet, it means we need to install it from its github repo by using the package `remotes`.

```{r, eval = FALSE}
remotes::install_github("tidyverse/multidplyr")
```

Since it is still under developement, use it with caution!


```{r}
library(tidyverse)
library(multidplyr)
```

There are two ways to get data to the workers in cluster:

- `partition()` a data frame that already loaded in the interactive process.
- Load a different subset of the data in each worker.


Use `partition()` to send data to workers
```{r}
library(nycflights13)

cluster <- new_cluster(4)

flights %>%
  group_by(dest, origin) %>%
  partition(cluster) %>%
  summarize(air_time = mean(air_time, na.rm = TRUE))
```


Load data in each worker
```{r}
cluster <- new_cluster(4)

cluster %>% cluster_library(c("tidyverse", "nycflights13"))

cluster %>%
  cluster_assign_partition(destination = unique(flights$dest))

# let's check the `destimation` variable in the workers
cluster %>%
  cluster_call(destination)

cluster %>% cluster_send({
  df <- flights %>% filter(dest %in% destination)
})

cluster %>%
  party_df("df") %>%
  group_by(dest, origin) %>%
  summarize(air_time = mean(air_time, na.rm = TRUE)) %>%
  collect()
```
