---
title: "Rcpp"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r, message = FALSE}
library(tidyverse)
```

# System requirements

To run c++ code in R, you will need to first install a workable toolchain.

- Windows
  - Install Rtools
    - R 3.6.x: https://cran.r-project.org/bin/windows/Rtools/history.html
    - R 4.0.x: https://cran.r-project.org/bin/windows/Rtools
- macOS
  - Run `which clang` in terminal to see if you have already installed it
  - If not, run `xcode-select --install` to install it.
- Linux
  - It depends

In R, we will need to install the package `Rcpp`.


The following sections will teach you the basics by translating simple R functions to their C++ equivalents

- No inputs, scalar output
- Scalar input and scalar output
- Vector input and scalar output
- Vector input and vector output
- Matrix input and vector output
- DataFrame input and DataFrame output


### No inputs, scalar output

This function has no arguments and always returns the integer 1:
```{r}
oneR <- function() 1L
```


```{Rcpp}
// [[Rcpp::export]]
int oneC() {
  return 1;
}
```


- You must declare the type of output the function returns. 
- Every statement is terminated by a `;`
- R doesn't have the idea of scalar variables but scalars and vectors in C++ are different. 

| R type           | Rcpp type       |
| -------          | ------          |
| int scalar       | int             |
| double scalar    | double          |
| logicial scalar  | bool            |
| character scalar | String          |
| int vector       | IntegerVector   |
| double vector    | NumericVector   |
| logicial vector  | LogicalVector   |
| character vector | CharacterVector |
  

```{r}
oneR()
oneC()
```

### Scalar input, scalar output

The next example function implements a scalar version of the `sign()` function which returns 1 if the input is positive, and -1 if it’s negative:

```{r}
signR <- function(x) {
  if (x > 0) {
    1L
  } else if (x == 0) {
    0L
  } else {
    -1L
  }
}
```


```{Rcpp}
// [[Rcpp::export]]
int signC(int x) {
  if (x > 0) {
    return 1;
  } else if (x == 0) {
    return 0;
  } else {
    return -1;
  }
}
```

```{r}
signC(0.8) # truncated to signC(0)
```


- We declare the type of each input in the same way we declare the type of the output. 
- The `if` syntax is identical. 
- The `while` loop syntax is almost the same, we use `break` to break the loop but use `continue` instead of `next` skip current iteration.

```{r}
while_exR <- function(x) {
  while(x < 9) {
    x <- x + 1
    if (x == 7) {
      break
    }
  }
  x
}
while_exR(1)
while_exR(7)
while_exR(10)
```


```{Rcpp}
// [[Rcpp::export]]
int while_exC(int x) {
  while(x < 9) {
    x = x + 1;
    if (x == 7) {
      break;
    }
  }
  return x;
}
```

```{r}
while_exC(1)
while_exC(7)
while_exC(10)
```

### Vector input, scalar output

It is where things get complicated.

```{r}
# it is the inefficient version of the function `sum`
sumR <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total
}
```


```{Rcpp}
// The first two lines import the namespace Rcpp for things such as `NumericVector`
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double sumC(NumericVector x) {
  int n = x.size();
  double total = 0;
  for (int i = 0; i < n; i++) {
    total += x[i];
  }
  return total;
}
```

- `NumericVector` is the Rcpp type of a double vector. To find its length, we could use the `.size()` method of `NumericVector`.
- We need to specify the type of each variable. `int n` and `double total`.
- The `for` loop starts with `i = 0` and ends if `i < n`. 
- After each iteration, we increment the value of `i` by one, `i++` increases the value of `i` by 1
- In C++, VECTOR INDICES START AT 0!!! The first element is `x[0]` and the last element is `x[n - 1]`;
- Use `=` for assignment, not `<-`
- `total += x[i]` means `total = total + x[i]`. Similarly, there are `-=`, `*=`, and `/=`.


Let's benchmark the performances
```{r, cached = TRUE, paged.print = FALSE}
x <- runif(1e3)
result <- bench::mark(
  sum(x),
  sumC(x),
  sumR(x),
  relative = TRUE
)
result
```
```{r}
autoplot(result, type = "violin")
```


Of course `sum` is the fastest because it is highly optimized.



### Vector input, vector output

This is a function that computes the Euclidean distance between a value and a vector of values:

```{r}
pdistR <- function(x, ys) {
  sqrt((x - ys) ^ 2)
}
```

```{r}
y <- runif(1e5)
# pdistR(0.5, y)
```

First, we translate the R function to a more naive R function based on a for loop.
```{r}
pdistR2 <- function(x, ys) {
  n <- length(ys)
  out <- numeric(length = n)
  for (i in seq_along(ys)) {
    out[i] = sqrt((x - ys[i]) ^ 2)
  }
  out
}
```


```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector pdistC(double x, NumericVector ys) {
  int n = ys.size();
  NumericVector out(n);

  for(int i = 0; i < n; i++) {
    out[i] = sqrt(pow(x - ys[i], 2.0));
  }
  return out;
}
```

- We create a new numeric vector of length `n` with a constructor: `NumericVector out(n);`. 
- Another useful way of making a vector is to copy an existing one: `NumericVector zs = clone(ys);`.
- C++ uses `pow()`, not `^`, for exponentiation.
- C++ is faster because it avoids the creation of temperary object like `x - ys` and `(x - ys)^2`.


```{r, paged.print = FALSE}
y <- runif(1e7)
result <- bench::mark(
  pdistR(0.5, y),
  pdistR2(0.5, y),
  pdistC(0.5, y)
)
result
```

- We saved about `1.6ms` by rewriting the R function in C++. Assume it takes 5 mins for you to write that function, you’d need to run it ~200 000 times to make rewriting worthwhile. 


### Sugar functions

Rcpp provides a number of sugar functions (check https://teuder.github.io/rcpp4everyone_en/210_rcpp_functions.html#list-of-r-like-functions for a list of such functions)
that make write C++ code easier.


```{Rcpp, eval = FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector slr_cpp(NumericVector x, NumericVector y) {
  double mux = mean(x);
  double muy = mean(y);
  double sxy = sum((x - mux) * (y - muy));
  double sxx = sum(pow(x - mux, 2));
  double slope = sxy / sxx;
  double intercept = muy - slope * mux;
  return NumericVector::create(intercept, slope);
}
```


#### Caution

Using sugar functions could be even slower than R functions.

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector pdistC_sugar(double x, NumericVector ys) {
  return sqrt(pow(ys - x, 2.0));
}
```

```{r, paged.print = FALSE}
y <- runif(1e6)
bench::mark(
  pdistR(0.5, y),
  pdistC(0.5, y),
  pdistC_sugar(0.5, y)
)
```

`pdistC_sugar` is the slowest among the three versions. The slowness is due to the creation of temporary objects such as `ys - x` and `pow(ys - x, 2.0)`.

### More on Vector Subsetting

Rcpp vectors could be extracted in "almost" the same way as R vectors.

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector positives(NumericVector x) {
    return x[x > 0];
}

// [[Rcpp::export]]
List first_three(List x) {
    IntegerVector idx = IntegerVector::create(0, 1, 2);
    return x[idx];
}

// [[Rcpp::export]]
List with_names(List x, CharacterVector y) {
    return x[y];
}
```

```{r}
x <- -5:5
positives(x)
l <- as.list(1:10)
first_three(l)
l <- l %>% set_names(letters[1:10])
with_names(l, c("a", "e", "g"))
```

Most excitingly, the subset mechanism works well with Rcpp sugar.

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector no_na(NumericVector x) {
    return x[!is_na(x)];
}
```

```{r, paged.print = FALSE}
R_no_na <- function(x) {
    return(x[!is.na(x)])
}
set.seed(1)
x <- rnorm(1e5)
x[sample(1e5, 1e4)] <- NA

bench::mark(
  R_no_na(x),
  no_na(x)
)
```

Of course, surgar functions could be slow. Instead, we could consider the function `R_IsNA` that checks if a number is NA.

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector no_na_fast(NumericVector x) {
  int n = x.size();
  int m = 0;
  int i, j = 0;
  for (i = 0; i < n; i++) {
    m += !R_IsNA(x[i]);
  }
  NumericVector y = NumericVector(m);
  for (i = 0; i < n; i++) {
    if (R_IsNA(x[i])) continue;
    y[j] = x[i];
    j++;
  }
  return y;
}
```

```{r}
autoplot(bench::mark(
  R_no_na(x),
  no_na(x),
  no_na_fast(x)
), type = "violin")
```

Again, we see that a naive for loop is faster.


### Values could be modified in place!!

```{r}
x <- 1:10
lobstr::obj_addr(x)
y <- x
lobstr::obj_addr(y)
y[2] <- 3
```

```{r}
lobstr::obj_addr(x)
```
```{r}
lobstr::obj_addr(y)
```
In R, values are not modified in place


Without cloning the object, any changes in C++ may modify the values in R.

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
void modifyVec(NumericVector x) {
  x[0] = 42;
}
```
PS: we use `void` because there are no return values.

```{r}
x <- c(3, 5)
lobstr::obj_addr(x)
```


```{r}
modifyVec(x)
x
```
```{r}
lobstr::obj_addr(x)
```


However, this side effect does not work if there is implict conversion.

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
void modifyVec2(IntegerVector x) {
  x[0] = 42;
}
```

```{r}
x <- c(3, 5)  # it is double
modifyVec2(x)  # a converted vector is passed to C++
x
```



### Matrix input and vector output


| R type | Rcpp type|
| -------| ------|
| int matrix | IntegerMatrix |
| double matrix | NumericMatrix |
| logicial matrix | LogicalMatrix |
| character matrix | CharacterMatrix |


Our next example is to compute column sum of a matrix. (This example is for demonstration only, in practice, we use `colSums` directly)

```{r}
colSumsR <- function(x) {
  m <- ncol(x)
  n <- nrow(x)
  output <- double(m)
  for (j in seq_len(m)) {
    output[j] = 0
    for (i in seq_len(n))
      output[j] = output[j] + x[i, j]
  }
  output
}
```


```{r}
(x <- matrix(rnorm(9), nc = 3))
colSumsR(x)
```

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector colSumsC(NumericMatrix x) {
  int m = x.ncol();
  int n = x.nrow();
  NumericVector output = NumericVector(m);
  for (int j = 0; j < m; j++) {
    output[j] = 0;
    for (int i = 0; i < n; i++)
      output[j] += x(i, j);
  }
  return output;
}

// [[Rcpp::export]]
NumericVector colSumsC_Sugar(NumericMatrix x) {
  return colSums(x);
}

```

- `.ncol()` and `.nrow()` return the number of columns and rows of the matrix
- we use `x(i, j)` instead of `x[i, j]` to retrieve the $i,j$ th element of the matrix

```{r}
x <- matrix(runif(1e6), 1e3)
autoplot(bench::mark(
  colSums(x),
  colSumsR(x),
  colSumsC(x),
  colSumsC_Sugar(x)
))
```


Again, we see that `colSumsC` is much faster than the pure R implementation `colSumsR`. The built-in `colSums` is the fastest because it is highly optimized.


### DataFrame input and DataFrame output


The following example modifies (not in place) some values of a dataframe.

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
DataFrame modifyDataFrame(DataFrame df) {
  // access the columns
  // it is very important to clone `a` and `b`!!
  // since `clone()` does not know what `df["a"]` is, we need to specify the target type
  NumericVector a = clone<NumericVector>(df["a"]);
  CharacterVector b = clone<CharacterVector>(df["b"]);
  // make some changes
  a[1] = 42;
  b[1] = "foo";
  // return a new data frame
  return DataFrame::create(_["a"]= a, _["b"]= b);
}
```

```{r}
df <- tibble(
    a = c(1, 2, 3),
    b = c("x", "y", "z"))
modifyDataFrame(df)
```

```{r}
df
```

The following example modifies (in place) some values of a dataframe.

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
void modifyDataFrame2(DataFrame df) {
  // casting the columns to their corresponding types
  NumericVector a = as<NumericVector>(df["a"]);
  CharacterVector b = as<CharacterVector>(df["b"]);
  // make some changes
  a[1] = 42;
  b[1] = "foo";
  // we don't have to return because we are modifying in place.
}
```

```{r}
df <- tibble(
    a = c(1, 2, 3),
    b = c("x", "y", "z"))
modifyDataFrame2(df)
df
```

#### To clone vs not to clone

- just want to read the vectors, use `as`
- want to modify values, use `clone` unless the modifications should be in place.


### List input

This is again not a good example because the same task could be easily done in R. It calcualtes the mean percentage error (`mpe()`) of a linear model.

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
```

```{r}
mpeR <- function(mod) {
  if (!inherits(mod, "lm")) {
    stop("Input must be a linear model")
  }
  mean(mod$residuals / (mod$fitted.values + mod$residuals))
}
```

```{r}
mpeR(mod)
```


```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double mpe(List mod) {
  if (!mod.inherits("lm")) stop("Input must be a linear model");

  // we don't have to clone because the values are not altered
  NumericVector resid = as<NumericVector>(mod["residuals"]);
  NumericVector fitted = as<NumericVector>(mod["fitted.values"]);

  int n = resid.size();
  double err = 0;
  for(int i = 0; i < n; ++i) {
    err += resid[i] / (fitted[i] + resid[i]);
  }
  return err / n;
}
```

```{r}
mpe(mod)
```

### Pass user function to Rcpp

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
RObject callWithTwo(Function f) {
  RObject f2 = f(2);
  // if you want to use named argument
  // return f(_["x"] = 2);
  return f2;
}

// [[Rcpp::export]]
NumericVector callWithTwoNumeric(Function f) {
  NumericVector f2 = as<NumericVector>(f(2));
  return f2;
}
```

PS: we use `RObject` because we don't know what it returns.

```{r}
f <- function(x) x + 1
g <- function(x) 3^x
callWithTwo(f)
callWithTwo(g)
```

```{r, error = TRUE}
callWithTwoNumeric(f)
h <- function(x) "hello"
callWithTwoNumeric(h)
```


## More advanced topics

- C++ Standard Template Library

  The standard template library (STL) provides a set of extremely useful data structures and algorithms. 
  However, it requires some background in C++ in order to disclose its full potential. (I will let you do your own reserach.)

- armadillo and eigen library

  If you want to do linear algebra opertions in C++, we should take a look at
  [Rcpparmadillo](https://CRAN.R-project.org/package=Rcpparmadillo) and   [RcppEigen](https://CRAN.R-project.org/package=RcppEigen).



# Reference

- Rcpp for everyone https://teuder.github.io/rcpp4everyone_en/
- Rcpp chapter of Advanced R https://adv-r.hadley.nz/rcpp.html
- Rcpp gallery https://gallery.rcpp.org/
- Rcpp Quick Reference Guide http://dirk.eddelbuettel.com/code/rcpp/Rcpp-quickref.pdf
