---
title: "Python interoperability"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: inline
---


# R vs Python

Why not both?


## Setup

### Step 1
To use python in R, we need to install the R package `reticulate`.
```{r}
library(reticulate)
```

### Step 2


Display python environment detected. You will be prompted to install `miniconda` if no suitable python is detected.

```{r, eval = FALSE}
py_discover_config()
```


In this class, we are going to focus on `miniconda.` If your system has another workable python environment, you will need to manually install `miniconda` with
```{r, eval = FALSE}
install_miniconda()
```
and specify to use `miniconda` after loading `reticulate`, like this

```{r, eval = FALSE}
library(reticulate)
use_miniconda(required = TRUE)  # sometimes, you may need to restart R
```

Visit [reticulate](https://rstudio.github.io/reticulate/articles/versions.html) website if you want to configure R to use different version of python then miniconda.


## Ways to use python

There are a variety of ways to integrate Python code into your R projects:

- Python in R Markdown


```{python}
import random
x = random.random()
x
```

```{r}
# to access Python objects in R
py$x
```

- Importing Python module

The `import()` function enables you to import any Python module and call its functions directly from R. 

```{r}
random <- import("random")
y <- random$random()
```

```{python}
# to access R object in python
r.y
```

The `import_builtins()` function enables to access the [built in functions](https://docs.python.org/3/library/functions.html#built-in-funcs).

```{r}
builtins <- import_builtins()
a <- builtins$range(5L)
builtins$len(a)
```


- Sourcing Python scripts: The `source_python()` function enables you to source a Python script the same way you would source() an R script (Unless `envir = NULL`, Python functions and objects defined within the script become directly available to the R session).


```{r}
source_python("script.py")
z
```


```{r}
# personally, I recommend using `envir = NULL` so no implicit copying is done
source_python("script.py", envir = NULL)
py$z
```


- Python REPL: `repl_python()`


## Python basics

https://learnxinyminutes.com/docs/python/
https://www.pyscoop.com/learn-python-in-10-minutes/


## Type conversions

https://rstudio.github.io/reticulate/#type-conversions

By default when Python objects are returned to R they are converted to their equivalent R types.

```{r}
random <- import("random")
(x <- random$random())
class(x)
```

However, if you'd rather make conversion from Python to R explicit and deal in native Python objects by default you can pass `convert = FALSE` to the import function

```{r}
random <- import("random", convert = FALSE)
(x <- random$random())
class(x)
```

We cannot work with native Python objects directly.
```{r, error = TRUE}
x + 1
```

The function `py_to_r` converts native Python objects to R objects.
```{r}
# convert x to R object first
py_to_r(x) + 1
```

Let's check another numpy example.
```{r}
np <- import("numpy", convert = FALSE)

# do some array manipulations with NumPy
a <- np$array(1:4)
s <- a$cumsum()

# convert to R explicitly at the end
py_to_r(s)
```

R Objects are converted to Python objects when they are passed to Python functions.

```{python}
def abssum(x):
    print("received a {} object\n".format(type(x)))
    return sum((abs(z) for z in x))
```

```{r}
x <- rnorm(10)
# x is implicitly to a native Python list
py$abssum(x)
# the input is converted to a python list explicitly
py$abssum(r_to_py(x))
```
```{r}
# there is no implicit conversion is the input is a python object
np <- import("numpy", convert = FALSE)
a <- np$array(1:4)
py$abssum(a)
```

## scikit-learn

[`scikit-learn`](https://scikit-learn.org/stable/modules/generated/sklearn.linear_model.LinearRegression.html) is a popular package for doing machine learning. It could be installed with
```{r, eval = FALSE}
py_install("scikit-learn")
```

```{r}
py_module_available("sklearn")
```

```{python}
import numpy as np
from sklearn.linear_model import LinearRegression
```

To use it,
```{python}
X = np.array([[1, 1], [1, 2], [2, 2], [2, 3]])
# y = 1 * x_0 + 2 * x_1 + e
beta = np.array([1, 2])
e = np.random.normal(0, 0.1, X.shape[0])
y = 3.14 + X.dot(beta) + e

lr = LinearRegression()
lr.fit(X, y)
print(lr.coef_)
print(lr.intercept_)
```

```{python}
Xtest = np.array([[1, 3]])
lr.predict(Xtest)
# double check
lr.intercept_ + Xtest.dot(lr.coef_)
```


How about passing R dataframes to it? Say we want to regress `mpg ~ hp * wt` of `mtcars`.

Method 1: prepare a function in python first

```{python}
from sklearn.linear_model import LinearRegression

def pylm(X, y):
  lr = LinearRegression()
  lr.fit(X, y)
  return lr
```

```{r}
fit <- lm(mpg ~ hp * wt, data = mtcars)
new_data <- data.frame(hp = 1.1, wt = 2.0)
predict(fit, new_data)
```

```{r}
# we do not need intercept
X <- model.matrix(~ -1 + hp * wt, mtcars)
y <- mtcars$mpg
fit <- py$pylm(X, y)
Xtest <- model.matrix(~ -1 + hp * wt, new_data)
fit$predict(Xtest)
```

Method 2: Import python module directly.

```{r}
sklearn <- import("sklearn")
LinearRegression <- sklearn$linear_model$LinearRegression

X <- model.matrix(~ -1 + hp * wt, mtcars)
y <- mtcars$mpg
lr <- LinearRegression()
lr$fit(X, y)

lr$coef_
```

```{r}
Xtest <- data.frame(hp = 1.1, wt = 2)
XMatrix <- model.matrix(~ -1 + hp * wt, Xtest)
lr$predict(XMatrix)
```
