---
title: "Functional programming in R"
output: 
  html_document:
    toc: yes
---

You'll learn some of the powerful programming tools provided by `purrr`, one of the tidyverse core packages.

```{r message=FALSE}
library(tidyverse)
```


## Pipe operator

We are going to detour a bit to talk about pipe operator.
Pipe operator is a powerful way to make your code much cleaner and readible by

- structuring sequences of data operations left-to-right (as opposed to from the inside and out),
- avoiding nested function calls,
- minimizing the need for local variables and function definitions, and
- making it easy to add steps anywhere in the sequence of operations.

The operators pipe their left-hand side values forward into expressions that
appear on the right-hand side, i.e. one can replace `f(x)` with 
`x %>% f()`, where `%>%` is the (main) pipe-operator. When coupling 
several function calls with the pipe-operator, the benefit will become
more apparent. Consider this pseudo example:


```{r, eval = FALSE}
raw_data <- read.csv("/path/to/data/file.csv")
sub_data <- subset(raw_data, variable_a > x)
new_data <- transform(sub_data, variable_c = variable_a / variable_b)
the_data <- head(new_data, 100)
```
However, the local variables are really not necessary, so one liner could be

```{r, eval = FALSE}
the_data <- head(
  transform(
    subset(
      read.csv("/path/to/data/file.csv"), 
    variable_a > x), 
  variable_c = variable_a / variable_b), 
100)
```

But that is too hard to read and reason. If we use pipe operator,

```{r, eval = FALSE}
the_data <- read.csv("/path/to/data/file.csv") %>%
  subset(variable_a > x) %>%
  transform(variable_c = variable_a / variable_b) %>%
  head(100)
```

Hint: In RStudio, you could use `Ctrl + Shift + M` (or `Cmd + Shift + M`) to insert the pipe operator.


### Basic usage

* `x %>% f`, `x %>% f()`, `x %>% f(.)`is equivalent to `f(x)`
* `x %>% f(y)` is equivalent to `f(x, y)`
* `x %>% f %>% g %>% h` is equivalent to `h(g(f(x)))`


```{r}
choose(5, 3)
5 %>% choose(3)
```


### The argument placeholder

* `x %>% f(y, .)` is equivalent to `f(y, x)`
* `x %>% f(y, z = .)` is equivalent to `f(y, z = x)`

```{r}
3 %>%  choose(5, .)
3 %>% choose(5, k = .)
# equivalent to 
choose(5, k = 3)
```

### Re-using the placeholder

It is straightforward to use the placeholder several times
in a right-hand side expression. However, when the placeholder
only appears in a nested expressions magrittr will still apply
the first-argument rule. The reason is that in most cases this
results more clean code. 

`x %>% f(y = nrow(.), z = ncol(.))` is equivalent to 
   `f(x, y = nrow(x), z = ncol(x))`

```{r}
5 %>% choose(. - 2)
choose(5, 5 - 3)
```

The behavior can be
overruled by enclosing the right-hand side in braces:

`x %>% {f(y = nrow(.), z = ncol(.))}` is equivalent to 
   `f(y = nrow(x), z = ncol(x))`
  
```{r}
5 %>% {choose(., . - 2)}
```


Piping two arguments
```{r}
list(n = 5, k = 3) %>% {choose(.$n, .$k)}
```


## What is functional programming?

> In computer science, functional programming is a programming paradigm - a style of building the structure and elements of computer programs that treats computation as the evaluation of mathematical functions and avoids changing-state and mutable data.

To put is simple, it is a programming style without side effects.


```{python, eval = FALSE}
def f(x):
  x["a"] = 2
  return x

x = {"a": 1}
f(x) == x
```

```{r}
f <- function(x) {
  x$a <- 2
  x
}

x <- list(a = 1)
isTRUE(all.equal(f(x), x))
```


## Is R a functional language?

R, at its heart, is a functional programming language. Most base R functions are pure, with a few notable exceptions:

- `library()` which loads a package, and hence modifies the search path.

- `setwd()`, `Sys.setenv()`, `Sys.setlocale()` which change the working directory, environment variables, and the locale, respectively.

- `plot()` and friends which produce graphical output.

- `write()`, `write.csv()`, `saveRDS()`, etc. which save output to disk.

- `options()` and `par()` which modify global settings.

- `S4` related functions which modify global tables of classes and methods.

- Random number generators which produce different numbers each time you run them

Almost everything else in R are pure functions.
```{r}
x <- 1
y <- 2
x + y
`+`(x, y)

a <- 3:5
a[3]
`[`(a, 3)

z <- 1
`<-`(z, 1)
```



## Revisit for loops

Suppose we want to compute the sample medians for each column of the data frame

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
# if you want to be fancy
df <- tibble(!!!rerun(4, rnorm(10)) %>% set_names(letters[seq_len(length(.))]))
```


A naive way to do so
```{r}
median(df$a)
median(df$b)
median(df$c)
median(df$d)
```

```{r}
# a base R for loop
output <- vector("double", ncol(df)) # 1. output
for (i in seq_along(df)) { # 2. sequence
  output[i] <- median(df[[i]]) # 3. body
}
output
```


### Base R solution

```{r}
lapply(df, median)
sapply(df, median)
```
These functions are great but they are not far less powerful then the next function `map`.


## Using `purrr`

THe `purrr` package (part of `tidyverse`) provides a function `map`.

The basic syntax for `map` is
```
map(ITERABLE, A_FUNCTION)
```
or
```
ITERABLE %>% map(A_FUNCTION)
```

* `ITERABLE`'s are objects like vectors, list and data frame.
* `A_FUNCTION` is a function, an expression, or a list of element names


```{r}
df %>% map(median)
```

Cosider the version of absolute which doesn't support vectors

```{r}
myabs <- function(x) {
  if (x < 0) {
    -x
  } else {
    x
  }
}
# myabs(c(-1, 3))  # error
c(-1, 3) %>% map_dbl(myabs)
```

Consider the following naive function which computes factorials
```{r}
fac <- function(n) {
  ## n has to be scalar
  if (n == 1) {
    return(1L)
  } else {
    return(fac(n - 1L) * n)
  }
}
fac(5)
```

Note that `fac` is not vectorized, it means that it doesn't handle vector input.
```{r, error = TRUE}
fac(1:5)
```

```{r}
1:5 %>% map(fac)
```
is almost equivalent to `list(fac(1), fac(2), fac(3), fac(4), fac(5))`


### Passing additional arguments

```r
mean(..., trim = 0.5)
```
computes trimmed mean of the input.

```{r}
df %>% map(mean, trim = 0.5)
list(
  a = mean(df$a, trim = 0.5), 
  b = mean(df$b, trim = 0.5), 
  c = mean(df$c, trim = 0.5), 
  d = mean(df$d, trim = 0.5))
```

### Inline function / expressions

```{r}
df %>% map(function(x) x^2)
```
or
```{r}
df %>% map(function(.) .^2)
```
or simply

```{r}
df %>% map_dfc(~ .^2)
```

### Name and position shortcuts

The `map` function also provide some shortcuts to extract elements from lists or vectors.


```{r}
# extract first element from each column
df %>% map(1)
```


```{r}
roster <- list(
  list(name = "Peter", id = 1231, grades = list(STA141A = "A", STA141B = "A"), hobbies = c("Math", "Stat")),
  list(name = "Mary",  id = 4213, grades = list(STA141A = "B", STA141B = "B"), hobbies = c("Math", "Chem"))
)
```

```{r}
roster %>% map("name")
```

### Multi-level extraction
```{r}
roster %>% map(list("grades", "STA141B"))
```
```{r}
roster %>% map(list("hobbies", 1))
```


### Type-specific map

`map()` always returns a list, even if all the elements have the same flavor and are of length one. But in that case, you might prefer a simpler object.

- `map_lgl()` makes a logical vector.
- `map_int()` makes an integer vector.
- `map_dbl()` makes a double vector.
- `map_chr()` makes a character vector.
- `mpa_dfr()` makes a data frame rowwise-ly
- `mpa_dfc()` makes a data frame columewise-ly

```{r}
df %>% map_dbl(mean)
```

```{r}
1:5 %>% map_int(fac)
```

```{r}
1:5 %>% map_lgl(~ . == 2L)
```

```{r}
roster %>% map_chr("name")
```


### Enumerating over the input

`imap` enumerates over the input

```{r}
a <- c(11, 12, 13)
a %>% imap_dbl(function(v, i) v)
```
```{r}
a %>% imap_dbl(function(v, i) i)
```

```{r}
a <- c(11, 12, 13)
b <- c(4, 5, 6)
a + b
a %>% imap_dbl(function(v, i) v + b[i])
# equivalently
a %>% imap_dbl(~ .x + b[.y])
```


### Extraction and assignment

```{r}
x <- 11:20
x %>% pluck(1) # equivalent to x[1]
```


Nested extraction
```{r}
a_to_e <- list(lowers = letters[1:5], uppers = LETTERS[1:5])
a_to_e %>% pluck("lowers") # equivalent to a_to_e["lowers"]
a_to_e %>% pluck("lowers", 2) # equivalent to a_to_e["lowers"][2]
```

Assignments
```{r}
a_to_e %>% assign_in(list("lowers", 2), "z")
# a_to_e["lowers"][2] <- "z"  # in-place modification
a_to_e %>% modify_in(list("lowers", 2), toupper)
```
These modifications are not in-place, the original variable is not modified
```{r}
a_to_e
```


### Keep and drop elements

```{r}
is_even <- function(x) x %% 2 == 0
11:20 %>% keep(is_even)
# or equivalently
11:20 %>% keep(~ . %% 2 == 0)
```

```{r}
11:20 %>% keep(~ !is_even(.))
11:20 %>% keep(negate(is_even))
11:20 %>% discard(is_even)
11:20 %>% discard(~ . %% 2 == 0)
```


### Predicate functions

```{r}
# is every element in `eleven_to_twenty` even?
11:20 %>% every(is_even) # more efficient then all(is_even(11:20))
# are some elements in `eleven_to_twenty` even?
11:20 %>% some(is_even) # more efficient then any(is_even(11:20))

11:20 %>% has_element(12L)
# caution! 12 is a double, but 11:20 is a integer vector.
11:20 %>% has_element(12)

11:20 %>% some(~ . == 12)  # maybe a bit robust

# the first number divisible by 7 is
11:20 %>% detect(~ . %% 7 == 0)
# get the index of the first element divisible by 7
11:20 %>% detect_index(~ . %% 7 == 0)
```

```{r}
# equivalently

x <- 11:20 
x %>% 
  detect_index(~ . %% 7 == 0) %>% 
  pluck(x, .)
```

### Map conditionaly

```{r}
square <- function(x) x^2
# square all even numbers
1:5 %>% map_if(is_even, square)
1:5 %>% map_if(~ . %% 2 == 0, ~ .^2)
```

```{r}
# apply the function to the third and fifth positions
1:5 %>% map_at(c(3, 5), square)
```

Note: these are currently no type specific `map_if` or `map_at`.

### Modify elements selectively

Unlike `map()` and its variants which always return a fixed object type, the `modify()` family always returns the same type as the input object.

```{r}
x <- rnorm(5)
x %>% modify(square)

# it doesn't work because 1:5 is an integer vector, while `square` produces double
# 1:5 %>% modify(square)
as.double(1:5) %>% modify(square)
1:5 %>% modify(~ as.integer(.^2))

add_one <- function(x) x + 1L
1:5 %>% modify(add_one) %>% class # integer
rnorm(5) %>% modify(add_one) %>% class # double

# make even numbers negative
1:5 %>% modify_if(is_even, ~ -.)

# make the third and fifth numbers negative
1:5 %>% modify_at(c(3, 5), ~ -.)
```
Again, these modifications are not in-place.

```{r}
x <- 1:5
x %>% modify_if(is_even, ~ -.)
x  # untouched
```

### Reshape lists

```{r}
a_to_e <- list(lowers = letters[1:5], uppers = LETTERS[1:5])
# a_to_e %>% flatten()
a_to_e %>% flatten_chr()  # type specific
```

```{r}
x <- matrix(1:6, 3, 2)
x
t(x)
```

```{r}
a_to_e %>%
  transpose() %>%
  map_chr(~ str_c(.$lowers, .$uppers))
```

```{r}
# without transpose, a little bit less readable
n <- length(a_to_e$lowers)
seq_len(n) %>% 
  map_chr(~ str_c(a_to_e$lowers[.], a_to_e$uppers[.]))
```


```{r}
mpgt <- mpg %>% transpose()

mpg$model[3]
mpgt[[3]]$model
```


```{r}
(M <- matrix(rnorm(9), nrow = 3, ncol = 3))

# base R
apply(M, 1, sd) # row sd's

# tidyverse version
M %>% 
  array_tree(1) %>% 
  map_dbl(sd)
```


### Add elements to lists

```{r}
li <- list(a = 1, b = 2)
# these aren't in-place modification
li %>%
  append(list(c = 3, d = 3)) %>%
  prepend(list(x = 0, y = 0)) %>%
  splice(u = 3, v = 4)
```

### Capture errors and warnings

```{r, error = TRUE}
log("a")  # gives error
log(-1)   # gives warning
```

```{r, error = TRUE}
list("a", -1, 3) %>% map(log)
```


```{r}
list("a", -1, 3) %>%
  map(safely(log)) %>% 
  map_dbl(~ if (is.null(.$error)) exp(.$result) else -1)
```


```{r}
list(-1, 3) %>%
  map(quietly(log)) %>% 
  map_dbl(~ if (length(.$warnings) == 0) exp(.$result) else -Inf)
```

```{r}
list("a", 3) %>%
  map_dbl(possibly(log, NaN))
```


### Map more than one argument

Two arguments
```{r}
a <- c(7, 8, 9)
b <- c(2, 3, 4)

map2_dbl(a, b, choose)
# list(choose(7, 2), choose(8, 3), choose(9, 4))
# seq_along(a) %>% map_dbl(~ choose(a[i], b[i]))

a %>% map2_dbl(b, choose)
b %>% map2_dbl(a, ., choose)
# or equivalently
map2_dbl(a, b, function(x, y) choose(x, y))
map2_dbl(a, b, ~ choose(.x, .y))
```

More than two arguments
```{r}
x <- list(
  a = c(1, 2, 3),
  b = c(5, 4, 8),
  c = c(4, 7, 9)
)

# x$a / (x$b + x$c)
x %>% pmap_dbl(function(a, b, c) a / (b + c))
```

### Cross

The `map` function applies to the inputs elementwisely. Sometimes, we wish to consider all combinations of the inputs. The `cross*` functions are designed to handle those situations.

`cross2` expects two iterables and usually is used together with `lift`
```{r}
a <- c(7, 8, 9)
b <- c(2, 3, 4)

cross2(a, b)[[1]] ## list(7, 2)
choose(7, 2)
li <- list(7, 2)
lifted_choose <- lift(choose)
lifted_choose(list(7, 2))
lift(choose)(list(7, 2))

cross2(a, b) %>% map_dbl(lift(choose))
```


`cross` expects a list of iterables
```{r}
li <- list(
  a = c(7, 8, 9),
  b = c(2, 3, 4)
)
li %>%
  set_names(c("n", "k")) %>%
  cross() %>%
  map_dbl(lift(choose))
```
The above operations do not keep track of the values of a and b. (PS: just a sidetrack, the following is not about functioning programming)

```{r}
data <- tibble(
  a = c(7, 8, 9),
  b = c(2, 3, 4)
)
```

```{r}
data %>% 
  cross_df() %>% 
  mutate(z = choose(a, b))  # assume the "function" is parallelizable
```

```{r}
data %>% 
  cross_df() %>% 
  rowwise() %>% 
  mutate(z = choose(a, b))  # assume the "function" is not parallelizable
```



```{r}
dat <- tibble(
  x = seq(-1, 1, length = 10),
  y = seq(-1, 1, length = 10)
)
dat
```

We wish to plot a heat map of `sqrt(x^2 + y^2)`
```{r}
dat %>% mutate(z  = sqrt(x^2 + y^2)) # wrong
```

```{r}
dat %>%
  cross_df() %>%
  mutate(z = sqrt(x^2 + y^2)) %>%
  ggplot(aes(x, y, fill = z)) +
  geom_tile()
```


```{r}
# alternatively
dat %>%
  expand(x, y) %>%
  mutate(z = sqrt(x^2 + y^2)) %>%
  ggplot(aes(x, y, fill = z)) +
  geom_tile()
```

The difference is that `expand` ignores duplication

```{r}
dat <- tibble(
  x = c(1, 1),
  y = c(2, 3)
)
dat %>% cross_df()
dat %>% expand(x, y)
```


### Reduce iterable

The `reduce` function combines the values of an iterable by iteratively applying a binary function.

```{r}
x <- 1:4
x %>% reduce(`+`)
# it means ((1 + 2) + 3) + 4
1:4 %>% reduce(`+`, .dir = "backward")
# it means 1 + (2 + (3 + 4))
```


It is usefully if we want to average data frames / matrices.

```{r}
matrices <- list(
  matrix(rnorm(9), nr = 3, nc = 3),
  matrix(rnorm(9), nr = 3, nc = 3),
  matrix(rnorm(9), nr = 3, nc = 3)
)

matrices %>% reduce(`+`) %>% `/`(length(matrices))
matrices %>% {reduce(., `+`)/ length(.)}
```


It is also useful if we wants to row bind。

```{r}
ll <- list(
  c(name = "Peter", age = 23),
  c(name = "Paul", age = 19),
  c(name = "Mary", age = 22)
)
```

```{r, error = TRUE}
bind_rows(ll)  #  it only works in dplyr 1.0
ll %>% reduce(bind_rows)
```


## A example of linear regression


```{r}
mtcars %>%
  nest(data = -cyl) %>% 
  mutate(model = map(data, ~ lm(mpg ~ wt, data = .))) %>%
  transmute(cyl, r_sq = model %>% map(summary) %>% map_dbl("r.squared"))
```


```{r}
# it is more readible
mtcars %>%
  group_by(cyl) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(mpg ~ wt, data = .))) %>%
  transmute(r_sq = model %>% map(summary) %>% map_dbl("r.squared"))
```

## Related concepts of R Functions

### Lexical scoping

```{r}
y <- 2
addy <- function(x) {
  x + y
}
addy(3)
```

```{r}
foo <- function(x) {
  addy(x)
}
foo(3)
```

`foo(3)` returns 3 because `addy` was defined in the global environment (the enclosing environment of `addy`).


The look up is dynamic though,
```{r}
y <- 4
addy(3)
```


### Dynamic scoping

Looking up variables in the calling environment rather than in the enclosing environment is called dynamic scoping. R is one of the few languages that implement dynamic scoping.

```{r}
y <- 2
addy <- function(x) {
  x + y
}

bar <- function(x) {
  y <- 4
  addy(x)
}
bar(3)
```

`bar(3)` returns 3 because of the lexical scroping rules of `addy`. However, we could change the scope rule by overriding the enclosing environment of a function.

```{r}
bar <- function(x) {
  y <- 4
  environment(addy) <- environment()
  addy(x)
}
bar(3)
```


### Lazy evaluation

In R, an expression is only evaluted when it is needed.

```{r}
one <- function() {
  cat("called one\n")
  1
}

square <- function(x) {
  cat("called square\n")
  x^2
}
square(one())
```

For comparsion, python is eager.
```{python}
def one():
  print("called one")
  return 1

def square(x):
  print("call square")
  return x**2
  
square(one())
```

R could do a lot of interesting things with lazy evaluation.

```{r}
echo <- function(x) {
  cat("you typed\n")
  print(substitute(x))
}

echo(it + does - not == have * to / make ^ sense)
```

### Some terminology

We end this chapter with several terminology.

- functionals
  - functions that have function inputs. For example, `map`: `map(x, mean)`.

- function factories
  - functions that return functions, For example,
    ```{r}
    pow <- function(k) {
      force(k)  # to avoid lazy evaluation
      function(x) {
        x^k
      }
    }
    square <- pow(2)
    square(5)
    ```

    ```{r}
    pow2 <- function(k) {
      function(x) {
        x^k
      }
    }
    h <- 2
    square <- pow2(h)
    h <- 3

    # up to this point, k or h is not evaluated
    square(2)  # 2^3

    h <- 4
    square(2)  # 2^3
    ```

- function operators
  - function that has function input(s) and returns a function. For examples, `lift`, `safely`.



# References

- R for Data Science https://r4ds.had.co.nz/iteration.html
- Advanced R https://adv-r.hadley.nz/fp.html
- purrr tutorial https://jennybc.github.io/purrr-tutorial/
- purrr cheatsheet https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf

