---
title: "Bag of little Bootstraps"
output: 
  html_document: 
    toc: yes
---


```{r}
library(tidyverse)
```

# Divide and conquer a.k.a. mapreduce does not always work

Divide and conquer allows a single task operation to be executed parallelly, but it does not always work.

```{r, echo = FALSE}
DiagrammeR::grViz("mapreduce.gv", height = 200)
```

Recall that there are two ways to get data to the workers in cluster:

- Partition a data set that already loaded in the main process.
- Load a different subset of the data in each worker.

The second appoarch is more efficient, so we first random split `flights` into 10 files.
In practice, the subsets are more likely to be loaded from a database server directly.

```{r}
library(nycflights13)
set.seed(141)
m <- 10
groups <- sample(seq_len(m), nrow(flights), replace = TRUE)
dir.create("flights/", showWarnings = FALSE)
for (i in seq_len(m)) {
  write_csv(filter(flights, groups == i), str_c("flights/", i, ".csv"))
}
```

As what we have done in Chapter 6. The mean could be computed by "mapreduce" with

```{r, message = FALSE}
file_names <- file.path("flights", list.files("flights"))
cor_list <- file_names %>%
  map_dbl(~ {
    df <- read_csv(., col_types = cols())
    cor(df$dep_delay, df$arr_delay, use = "pairwise.complete.obs")
  })
(mean_cor <- cor_list %>% reduce(`+`) / m)
```

```{r}
cor(flights$dep_delay, flights$arr_delay, use = "pairwise.complete.obs")
```


You may wonder if you could do the same for confidence intervals.
```{r, message = FALSE}
ci_list <- file_names %>%
  map(~ {
    df <- read_csv(., col_types = cols())
    cor.test(df$dep_delay, df$arr_delay)$conf.int
  })
(mean_ci <- ci_list %>% reduce(`+`) / m)
```
Yeah, it gives us a result. But wait, it doesn't look right. Though the mapreduce procedure speeds up the computation, it should give similar result as if we work on the whole dataset.

```{r}
cor.test(flights$dep_delay, flights$arr_delay)$conf.int
```


*Lesson learned*: we cannot combine any statistics in the reduce step by simply taking the average. We may need to scale the statistics analytically which could be hard or impossible.


# The bag of little bootstraps (BLB)

It is a procedure which incorporates features of both the bootstrap and subsampling to yield a robust, computationally efficient means of assessing the quality of estimators


```{r, echo = FALSE}
DiagrammeR::grViz("blb.gv", height = 300)
```


Bascially, the bag of little bootstraps = subsample + bootstrap. However, for each bootstrap, we sample $n$ from $b$ with replacement instead of sample $b$ from $b$ as in oridinary bootstrap.

- sample without replacement the sample $s$ times into sizes of $b$
- for each subsample
  - resample each until sample size is $n$, $r$ times
  - compute the bootstrap statistic (e,g., the mean of a variable, or cor between two variables) for each bootstrap sample
  - compute the statistic (e.g., confidence interval) from the bootstrap statistics
- take the average of the statistics


## A naive (single core) implementation

```{r, message = FALSE}
r <- 10 # r should be at least a few thousands, say 10000, we are using 10 for demo
n <- nrow(flights)

ci_list <- file_names %>% map(~ {
  df <- read_csv(., col_types = cols()) %>%
    select(dep_delay, arr_delay)
  seq_len(r) %>%
    map_dbl(~ {
      df <- df[sample(seq_len(nrow(df)), n, replace = TRUE), ]
      cor(df$dep_delay, df$arr_delay, use = "pairwise.complete.obs")
    }) %>%
    quantile(c(0.025, 0.975))
})

reduce(ci_list, `+`) / length(ci_list)
```

The result is much closer to the result from the `cor.test` based on the whole data. (even `r` was set as 10)

However, the above implementation is not memory and computationally efficient because some rows in `df` are duplicated.


## A more efficient implmentation with multinomial distribution

We could use multinomial distribution to represent the frequency of each row.

Imagine that we have a dataset of original size of 100 and each subsample is of size 10
```{r}
n <- 100
df <- tibble(x = rnorm(n), y = rnorm(n))
df1 <- df %>% slice(1:10)
```

We will need sample `n` obs from `df1` which replacement. Naively,
```{r}
df1[sample(seq_len(nrow(df1)), n, replace = TRUE), ] %>%
  arrange(x, y)
```

But a lot of rows are duplicated!

A more efficient way is to first generate the repetitions by multinomial distribution.

```{r}
df1 %>%
  mutate(freq = rmultinom(1, n, rep(1, n())))
```

*Compute the statistic with the frequencies*

```{r, message = FALSE}
n <- nrow(flights)
df <- read_csv(file_names[1], col_types = cols()) %>%
  select(dep_delay, arr_delay) %>%
  drop_na() %>%
  mutate(freq = rmultinom(1, n, rep(1, n())))
df
```

How do you calculate the mean based on frequencies? (weighted mean!)
```{r}
with(df, weighted.mean(dep_delay, freq))
```
```{r}
df %>%
  mutate(s = dep_delay * freq) %>%
  summarize(sum(s) / sum(freq)) %>%
  pull()
```


What if we want weighted correlation between two variables? Base R doesn't provide function to compute weighted correlations. Let's do it manually.

Recall that ordinary correlation is
$$
r = \frac{\sum (x_i-\bar x) (y_i-\bar y)}{\sqrt{\sum (x_i-\bar x)^2  \sum (y_i-\bar y)^2 }}
$$
But now some $(x_i, y_i)$ are repeated, to be more precise, each of them repeated $w_i$ times. So the formula is now
$$
r = \frac{\sum w_i (x_i-\bar x_w) (y_i-\bar y_w)}{\sqrt{\sum w_i (x_i-\bar x_w)^2  \sum w_i (y_i-\bar y_w)^2 }}
$$
where $\bar x_w = \sum(w_i x_i)/ n$, $\bar y_w = \sum(w_i y_i)/ n$ and $n = \sum w_i$

Of course, it is not computationally efficient, with some algebra, the computational formula is derived

$$
r = \frac{
  \sum w_i x_i y_i-  (\sum w_i x_i \sum w_i y_i)/n
}{
  \sqrt{\left[\sum w_i x_i^2- (\sum w_i x_i)/n \right]\left[ \sum w_i y_i^2-  (\sum w_i y_i)/n \right] }
}
$$
where $n = \sum w_i$.

```{r}
weighted_cor <- function(x, y, weights) {
  n <- sum(weights)
  sumx <- sum(weights * x)
  sumy <- sum(weights * y)
  sxx <- sum(weights * x^2) - sumx^2 / n
  syy <- sum(weights * y^2) - sumy^2 / n
  sxy <- sum(weights * x * y) - sumx * sumy / n
  sxy / sqrt(sxx * syy)
}
```
PS: there are still ways to improve it.


Compare it with the duplicated data to make sure that `weighted_cor` works
```{r}
df %>%
  slice(rep(seq_len(n()), freq)) %>%
  summarise(cor(dep_delay, arr_delay))
```
```{r}
with(df, weighted_cor(dep_delay, arr_delay, freq))
```


*Put everything back*

```{r, message = FALSE}
r <- 100 # r should be at least a few thousands, we are using 100 for demo
n <- nrow(flights)


ci_list <- file_names %>% map(~ {
  df <- read_csv(., col_types = cols()) %>%
    select(dep_delay, arr_delay) %>%
    drop_na()
  map_dbl(seq_len(r), ~ {
    freq <- rmultinom(1, n, rep(1, nrow(df)))
    with(df, weighted_cor(dep_delay, arr_delay, freq))
  }) %>%
    quantile(c(0.025, 0.975))
})

reduce(ci_list, `+`) / length(ci_list)
```


## A parallel version using `furrr`.


```{r, message = FALSE}
library(furrr)
plan(multiprocess, workers = 4)
```

```{r, message = FALSE}
ci_list <- file_names %>% future_map(~ {
  df <- read_csv(., col_types = cols()) %>%
    select(dep_delay, arr_delay) %>%
    drop_na()
  map_dbl(seq_len(r), ~ {
    freq <- rmultinom(1, n, rep(1, nrow(df)))
    unclass(with(df, weighted_cor(dep_delay, arr_delay, freq)))
  }) %>%
    quantile(c(0.025, 0.975))
})

reduce(ci_list, `+`) / length(ci_list)
```
