---
title: "Debugging in R"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---


# Debugging

Jenny Bryan gave a very good talk in [RStudio conf 2020](https://resources.rstudio.com/rstudio-conf-2020/object-of-type-closure-is-not-subsettable-jenny-bryan).

- reset<br>
  Jenny recommends [not saving .Rdata](https://speakerdeck.com/jennybc/object-of-type-closure-is-not-subsettable?slide=12) on exit.
- reprex<br>
  `reprex` is a package which makes reproducible examples.
- debug<br>
  the main focus of this chapter
- deter<br>
  Fix it once? Keep it fixed.


## Different types of errors

- syntax / parsing error
- runtime error
- logical error


## Locating errors

```{r, error = TRUE}
f <- function(a) g(a)
g <- function(b) h(b)
h <- function(c) i(c)
i <- function(d) {
  if (!is.numeric(d)) {
    stop("`d` must be numeric", call. = FALSE)
  }
  d + 10
  
}
f("a")
```

See also `debug.R` demonstration.


## Lazy evaluation

```{r, error = TRUE}
j <- function() k()
k <- function() stop("Oops!", call. = FALSE)
f(j())
```
```{r, error = TRUE}
rlang::with_abort(f(j()))
rlang::last_trace()
```

## Using `recover()`

By setting ,
```{r, eval = FALSE}
original_error_handler = getOption("error")
options(error = recover)
```
a interactive prompt will be displayed that you get an error.

```{r, eval = FALSE}
# restore the original error handler
options(error = original_error_handler)
```


## Interactive debugger in RStudio

When we encounter an error, we could hit `Rerun with Debug` in RStudio to start interactive debugging. 

```{r, error = TRUE}
f("a")
```

Sometimes, we know something is definitely wrong but the code runs fine. 

## Interactive debugger elsewhere

### `browser()`

When `browser()` is run, a interactive prompt will be shown.

```{r}
g <- function(b) {
  browser()
  h(b)
}
f("a")
```

`browser()` is a regular function call which means that you can run it conditionally

```{r, eval = FALSE}
g <- function(b) {
  if (b < 0) {
    browser()
  }
  h(b)
}
g(10)
```

See `debug2.R` for demonstration.

### `debug()` and `debugonce()`


`debug` takes a single argument, the name of a function. When you pass the name of
a function to debug, that function is flagged for debugging.

See `debug3.R` for demonstration.

See `debug4.R` for demonstration.


# Reference

Advanced R Chapter 22 https://adv-r.hadley.nz/debugging.html
